---
title: "Title"
author: 
- name: "Regina H. Reynolds"
  affiliation: UCL
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  bookdown::html_document2:
    figure_caption: yes
    code_folding: show
    theme: paper
    highlight: kate
    df_print: paged
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include = FALSE}

library(ggplot2) # For plotting
library(tidyverse) # For tidy manipulation of data
library(stringr) # For string manipulation

knitr::opts_chunk$set(echo = T, warning = F, message = F)

# Set defaults for ggplots 
theme_rhr <- theme_set(
  theme_bw(base_family = "Helvetica",
           base_size = 10) + 
  theme(panel.grid.major.x = element_blank(),
        legend.position = "top",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.title.y = element_text(vjust = 0.6),
        panel.spacing = unit(0.1, "lines"))
)

```

> Aim: What's the major aim of this .Rmd?
<br><br>

# Background

What is the background for running this analysis?

# Methods {.tabset}

Describe methods in text, as you would do in a paper.

# Supplementary code {.tabset}

Following section includes any intermediary code used in this `.Rmd`.

## Example supplementary code section

```{r example-supp-code, eval = F}

source(here::here("", ""))

```

# Results 

## Example results section {.tabset}

### Text

### Figures

For figures, need to make sure your keys don't have a "_" in them and have `fig.cap = "figure caption"` to refer to them in the text (\@ref(fig:iris-figure))

```{r iris-figure, fig.cap = "You need this to refer to figures"}

iris_10 <- data.frame(iris[1:10,]) 

iris_10 %>% 
    ggplot(aes(x = Sepal.Length,
               y = Sepal.Width)
    ) +
    geom_point()

```

<br><br>

# Session info

```{r reproducibility, echo = FALSE}
# Session info
library("sessioninfo")
options(width = 120)
session_info()
```
